Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile

### 2024-01-24 article/paper_drought_Chile2023.tex
actualizar discusion

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p,
  onecolumn]{elsarticle}

@@

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

The primary cause of drought is precipitation, and temperature makes it
worse \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citet{Loon2016} and
\citet{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought, \citet{Crausbay2017}
proposed the ecological drought definition as ``an episodic deficit in
water availability that drives ecosystems beyond thresholds of
vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems.''. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} predicts that many regions of the
world will experience more severe agricultural and ecological droughts
even if global warming stabilizes at 1.5°--2°C. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

@@

To evaluate meteorological drought (i.e., water supply), the World
Meteorological Organization (WMO; \citet{WMO2012}) recommends the
Standardized Precipitation Index (SPI; \citet{Mckee1993}), a multiscalar
drought index that allows to monitor precipitation deficits from short-
to long-term. Following the same approach, \citet{Vicente-Serrano2010}
incorporates into the SPI the effect of temperature through the use of
potential evapotranspiration, thus proposing the SPEI (Standardized
Precipitation Evapotranspiration Index). Similarly, to evaluate solely
the evaporative demand driven by temperature, \citet{Hobbins2016} and
\citet{McEvoy2016} came up with the Evaporative Demand Drought Index
(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
\citet{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
cumulative Normalized Difference Vegetation Index (NDVI), which could be
acumulated over the growing season or any period (e.g., months),
resulting in a multiscalar drought index. For soil moisture, several
drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
normalized index \citep{Narasimhan2005} and the Soil Moisture
Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
normalized index using vegetation, land surface temperature, and a
vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
estimate the standardized terrestrial water storage index (STI)
\citep{Cui2021}, a standardized anomaly that follows the methodology of
the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
water supply, demand, and storage, which can help to make a
comprehensive assessment of drought.

In this research, we aim to analyze the impact of drought on different
types of ecosystems (land cover classes) in continental Chile. Our
specific goals are: i) to analyze the trend on multi-scalar drought
indices for water demand and supply, soil moisture, and vegetation
productivity for 1981--2023/2001--2023; ii) to assess the LULC change
for 2001--2021 and how it relates to drought indices; iii) to evaluate
the relationship between zcNDVI, a measure of vegetation productivity,
and drought indices for water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citet{Beck2023} and the land cover
type persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
There is a significant amount (50\%) of Chilean matorral (shrubland and
savanna, \citep{Fuentes2021}), then grassland (16\%), forest (8\%), and
croplands (5\%), in ``Centro.'' The south part of ``Sur'' and the north
part of ``Austral'' are dominated by an oceanic climate (Cfb). Those
zones are high in forest and grassland. The southern part of the country
has a tundra climate, and in Patagonia, it is a cold semi-arid area with
an extended surface of grassland, forest, and, to a lesser extent,
savanna.

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
Before using the NDVI, it was smoothed using a locally-weighted
polynomial regression, following the procedure described in
\citet{Zambrano2018} and \citet{Zambrano2016}.

@@

It corresponds to a moving window (convolution) that sums the variable
for \(s\) starting for the last month \(n\) until the month, which could
sum for \(s\) months (n-s+1). Once the variable is cumulated over time
for the scale \(s\), we used a nonparametric approach following
\citet{Hobbins2016} to derive the drought indices. Thus, the empirically
derived probabilities are obtained through an inverse normal
approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
plotting position \citep{Wilks2011} over \(A_i\) to derive the
\(P(A_i)\) probabilities across a period of interest:

@@

\hypertarget{land-cover-macroclasess-and-validation}{%
\subsubsection{land cover macroclasess and
validation}\label{land-cover-macroclasess-and-validation}}

@@

To validate the land cover obtained, we compare the macroclasses with
the ones of a more detailed land cover map made by \citet{Zhao2016} for
Chile with samples acquired in the years 2013--2014 (LCChile). The later
has a spatial resolution of 30 m and three levels of defined classes;
from those, we used level 1, which fits with the macroclasses land
cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land
cover macrolcasses to validate with LCChile.

@@

\hypertarget{land-cover-persistence-mask-2001-2022}{%
\subsubsection{land cover persistence mask
2001-2022}\label{land-cover-persistence-mask-2001-2022}}

@@

\hypertarget{land-cover-trend-and-drought-indices}{%
\subsubsection{land cover trend and drought
indices}\label{land-cover-trend-and-drought-indices}}

@@

\citet{Vicente-Serrano2021} made a global analysis of the drought's
severity trend using SPI, SPEI, and the Standardized Evapotranspiration
Deficit Index (SEDI; \citet{Vicente-Serrano2018}) to evaluate AED. They
indicate that the increase in hydrological drought has been due to
anthropogenic effects rather than climate change. This is because the
global increase in AED did not explain the change in the spatial pattern
of the hydrological drought. Also, they state that \emph{``the increase
in hydrological droughts has been primarily observed in regions with
high water demand and land cover change''}. We will contrast this
hypothesis with what is occurring in Chile. To achieve this, we will use
the land cover class type that remains more than 80\% of types for
2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
there are low changes.

@@

To assess how water demand and supply and soil moisture are related to
vegetation productivity (zcNDVI), we analyze the linear correlation
between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and
36-month time scales against zcNDVI. We followed a similar approach to
that used by \citet{Meroni2016} when using the SPI for meteorological
drought against the cumulative FAPAR (Fraction of Absorbed
Photosynthetically Active Radiation) as a proxy for vegetation
productivity. We made a pixel-to-pixel linear correlation analysis per
index. First, we calculate the Pearson coefficient of correlation for
the six time scales and let the time scale that reaches the maximum
correlation be significant (p \textless{} 0.05). Then, we extracted the
Pearson correlation value corresponding to the time scales that reached
the maximum value. Thus, we derived two raster maps per index, the first
with the time scales and the second with the correlation value.

@@

\hypertarget{land-cover-macroclasess-and-validation-1}{%
\subsubsection{land cover macroclasess and
validation}\label{land-cover-macroclasess-and-validation-1}}

@@

\hypertarget{land-cover-persistence-mask-2001-2022-1}{%
\subsubsection{land cover persistence mask
2001-2022}\label{land-cover-persistence-mask-2001-2022-1}}

@@

\hypertarget{land-cover-trend-and-drought-indices-1}{%
\subsubsection{land cover trend and drought
indices}\label{land-cover-trend-and-drought-indices-1}}

@@

Regarding vegetation productivity aggregated through the macrozones in
the five land cover macroclasses, in ``Norte Grande,'' there is an
increase trend of 0.02 (z-index) per decade, related to types of
grassland and shrubland. There is a negative trend in ``Norte Chico''
with -0.04 and ``Centro'' with -0.02 per decade. In the ``Norte Chico,''
savanna (-0.05) has the lowest trend, and the rest of the types are
around -0.04. In ``Centro,'' shrubland reaches -0.06, grassland -0.05,
and croplands and savanna -0.01 per decade. This could be associated
either with a reduction in vegetation surface, a decrease in biomass, or
browning \citep{Miranda2023}. Vegetation reached its lowest values since
the year 2019, reaching an extreme condition in early 2020 and 2022 in
the ``Norte Chico'' and Centro'' (Mega Drought). The ``Sur'' and
``Austral'' show a positive trend of around 0.016 per decade
(Figure~\ref{fig-zcNDVI_var}). Despite the croplands suffering from
drought just as badly as the native vegetation in ``Norte Chico,'' the
Chilean matorral appears to be the region most affected by a negative
trend in vegetation \citep{Fuentes2021}.

@@

\hypertarget{drought-trend-lulcc-and-climate-conditions}{%
\subsection{Drought trend, LULCC, and climate
conditions}\label{drought-trend-lulcc-and-climate-conditions}}

1.- Respecto a lo que indica \citet{Vicente-Serrano2018}, de que el
aumento en la tendencia en severidad de la sequía (hidrológica) tiene
que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas)
que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en
Chile?

\hypertarget{land-cover-types-most-impacted-by-drought-throughout-chile}{%
\subsection{land cover types most impacted by drought throughout
Chile}\label{land-cover-types-most-impacted-by-drought-throughout-chile}}

2.- Sobre los tipos de land cover más afectados por los indicadores de
sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur
(\citet{Miranda2023})

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  preprint,
  3p,
  onecolumn]{elsarticle}

@@

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

The primary cause of drought is precipitation, and temperature makes it
worse \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citep{Loon2016} and
\citep{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought, \citep{Crausbay2017}
proposed the ecological drought definition as ``an episodic deficit in
water availability that drives ecosystems beyond thresholds of
vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems.''. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} predicts that many regions of the
world will experience more severe agricultural and ecological droughts
even if global warming stabilizes at 1.5°--2°C. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

@@

To evaluate meteorological drought (i.e., water supply), the World
Meteorological Organization (WMO; \citep{WMO2012}) recommends the
Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
drought index that allows to monitor precipitation deficits from short-
to long-term. Following the same approach, \citep{Vicente-Serrano2010}
incorporates into the SPI the effect of temperature through the use of
potential evapotranspiration, thus proposing the SPEI (Standardized
Precipitation Evapotranspiration Index). Similarly, to evaluate solely
the evaporative demand driven by temperature, \citep{Hobbins2016} and
\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
cumulative Normalized Difference Vegetation Index (NDVI), which could be
acumulated over the growing season or any period (e.g., months),
resulting in a multiscalar drought index. For soil moisture, several
drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
normalized index \citep{Narasimhan2005} and the Soil Moisture
Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
normalized index using vegetation, land surface temperature, and a
vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
estimate the standardized terrestrial water storage index (STI)
\citep{Cui2021}, a standardized anomaly that follows the methodology of
the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
water supply, demand, and storage, which can help to make a
comprehensive assessment of drought.

In this research, we aim to analyze the impact of drought on different
types of ecosystems (land cover classes) in continental Chile. Our
specific goals are: i) to analyze the trend on multi-scalar drought
indices for water demand and supply, soil moisture, and vegetation
productivity for 1981--2023/2001--2023; ii) to assess the LULC change
for 2001--2022 and how it relates to drought indices; iii) to evaluate
the relationship between zcNDVI, a measure of vegetation productivity,
and drought indices for water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citep{Beck2023} and the land cover
type persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
There is a significant amount (50\%) of Chilean matorral (shrubland and
savanna, \citep{Fuentes2021}), then grassland (16\%), forest (8\%), and
croplands (5\%), in ``Centro.'' The south part of ``Sur'' and the north
part of ``Austral'' are dominated by an oceanic climate (Cfb). Those
zones are high in forest and grassland. The southern part of the country
has a tundra climate, and in Patagonia, it is a cold semi-arid area with
an extended surface of grassland, forest, and, to a lesser extent,
savanna.

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
Before using the NDVI, it was smoothed using a locally-weighted
polynomial regression, following the procedure described in
\citep{Zambrano2018} and \citep{Zambrano2016}.

@@

It corresponds to a moving window (convolution) that sums the variable
for \(s\) starting for the last month \(n\) until the month, which could
sum for \(s\) months (n-s+1). Once the variable is cumulated over time
for the scale \(s\), we used a nonparametric approach following
\citep{Hobbins2016} to derive the drought indices. Thus, the empirically
derived probabilities are obtained through an inverse normal
approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
plotting position \citep{Wilks2011} over \(A_i\) to derive the
\(P(A_i)\) probabilities across a period of interest:

@@

\hypertarget{land-cover-macroclasess-and-validation}{%
\subsubsection{Land cover macroclasess and
validation}\label{land-cover-macroclasess-and-validation}}

@@

To validate the land cover obtained, we compare the macroclasses with
the ones of a more detailed land cover map made by \citep{Zhao2016} for
Chile with samples acquired in the years 2013--2014 (LCChile). The later
has a spatial resolution of 30 m and three levels of defined classes;
from those, we used level 1, which fits with the macroclasses land
cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land
cover macrolcasses to validate with LCChile.

@@

\hypertarget{land-cover-persistence-mask-2001-2022}{%
\subsubsection{Land cover persistence mask
2001-2022}\label{land-cover-persistence-mask-2001-2022}}

@@

\hypertarget{land-cover-trend-and-drought-indices}{%
\subsubsection{Land cover trend and drought
indices}\label{land-cover-trend-and-drought-indices}}

@@

\citep{Vicente-Serrano2021} made a global analysis of the drought's
severity trend using SPI, SPEI, and the Standardized Evapotranspiration
Deficit Index (SEDI; \citep{Vicente-Serrano2018}) to evaluate AED. They
indicate that the increase in hydrological drought has been due to
anthropogenic effects rather than climate change. This is because the
global increase in AED did not explain the change in the spatial pattern
of the hydrological drought. Also, they state that \emph{``the increase
in hydrological droughts has been primarily observed in regions with
high water demand and land cover change''}. We will contrast this
hypothesis with what is occurring in Chile. To achieve this, we will use
the land cover class type that remains more than 80\% of types for
2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
there are low changes.

@@

To assess how water demand and supply and soil moisture are related to
vegetation productivity (zcNDVI), we analyze the linear correlation
between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and
36-month time scales against zcNDVI. We followed a similar approach to
that used by \citep{Meroni2016} when using the SPI for meteorological
drought against the cumulative FAPAR (Fraction of Absorbed
Photosynthetically Active Radiation) as a proxy for vegetation
productivity. We made a pixel-to-pixel linear correlation analysis per
index. First, we calculate the Pearson coefficient of correlation for
the six time scales and let the time scale that reaches the maximum
correlation be significant (p \textless{} 0.05). Then, we extracted the
Pearson correlation value corresponding to the time scales that reached
the maximum value. Thus, we derived two raster maps per index, the first
with the time scales and the second with the correlation value.

@@

\hypertarget{land-cover-macroclasess-and-validation-1}{%
\subsubsection{Land cover macroclasess and
validation}\label{land-cover-macroclasess-and-validation-1}}

@@

\hypertarget{sec-persistence_mask}{%
\subsubsection{Land cover persistence mask
2001-2022}\label{sec-persistence_mask}}

@@

\hypertarget{land-cover-trend-and-drought-indices-1}{%
\subsubsection{Land cover trend and drought
indices}\label{land-cover-trend-and-drought-indices-1}}

@@

Regarding vegetation productivity aggregated through the macrozones in
the five land cover macroclasses, in ``Norte Grande,'' there is an
increase trend of 0.02 (z-index) per decade, related to types of
grassland and shrubland. There is a negative trend in ``Norte Chico''
with -0.04 and ``Centro'' with -0.02 per decade. In the ``Norte Chico,''
savanna (-0.05) has the lowest trend, and the rest of the types are
around -0.04. In ``Centro,'' shrubland reaches -0.06, grassland -0.05,
and croplands and savanna -0.01 per decade. This could be associated
either with a reduction in vegetation surface, a decrease in biomass, or
browning \citep{Miranda2023}. Vegetation reached its lowest values since
the year 2019, reaching an extreme condition in early 2020 and 2022 in
the ``Norte Chico'' and Centro'' (Mega Drought). The ``Sur'' and
``Austral'' show a positive trend of around 0.016 per decade
(Figure~\ref{fig-zcNDVI_var}). Despite the croplands suffering from
drought just as badly as the native vegetation in ``Norte Chico,'' the
Chilean matorral \citep{Fuentes2021} appears to be the region most
affected by a negative trend in vegetation productivity.

@@

\hypertarget{drought-trend-and-attribution-to-lulcc}{%
\subsection{Drought trend and attribution to
LULCC}\label{drought-trend-and-attribution-to-lulcc}}

\citep{Vicente-Serrano2021}, in a study at the global scale of drought
trends, indicates that there have not been significant trends in
meteorological drought since 1950. Also, state that the increase in
hidrological trend in some parts of the globe (northeast Brazil and the
Mediterranean region) is related to changes in land cover and
specifically to the rapidly increasing irrigated area, which
consequently increases water extraction. \citep{Kogan2020} analyzed the
agricultural drought impact globally and in the main grain producer
countries, finding that \emph{``since 1980, the Earth warming has not
changed the drought area or intensity''}. In our study, we considered
the variation in vegetation productivity in Chile for areas without
changes in land cover macroclasses (see
Section~\ref{sec-persistence_mask}), to avoid misleading conclusions
that could be related to the increase in water demand due to LULCC. Our
results show a contrasting perspective. There has been a significant
trend in the decline of vegetation productivity (zcNDVI) since 2000 for
``Norte Chico'' and ``Centro,'' which has been extreme between 2020 and
2022, seemsly due to an intense hydrological drought due to the
persistance of the Mega Drought \citep{Garreaud2017}. Despite we used
the persistance mask for vegetations's trend analysis, the cropland,
which is the most water-demand type, showed a decrease trend in ``Norte
Chico'' and ``Centro.'' Also, there was an increase in barren land for
both types. These changes are associated with a decrease in water demand
from vegetation. On the other hand, for ``Norte Chico'' and ``Centro,''
our results show a decrease in trends of water supply (SPI and SSI),
which are higher at larger time scales and consequently impact the
hydrological system. We claim that what occurred in central Chile defies
findings made at the global level
\citep{Vicente-Serrano2021, Kogan2020}, demonstrating that a constant
decrease in water supply rather than an increase in water demand (i.e.,
irrigated crops) is the main cause of the hydrological drought. Finally,
central Chile has shown a diminishment in vegetation productivity for
all macroclasses, mainly attributed to variation in water supply, i.e.,
precipitation, which could be strengthened by an increase in water
demand by, for example, irrigated crops.

1.- Respecto a lo que indica \citep{Vicente-Serrano2022}, de que el
aumento en la tendencia en severidad de la sequía (hidrológica) tiene
que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas)
que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en
Chile?

\citep{Chen2022}: Assessing the impact of drought-land cover change on
global vegetation greenness and productivity

\citep{Kogan2020}: muestra que la sequía no se ha intensificado ni
expandido a nivel global, por 38 años.

\hypertarget{land-cover-types-and-their-impact-by-drought}{%
\subsection{Land cover types and their impact by
drought}\label{land-cover-types-and-their-impact-by-drought}}

Extreme drought conditions are an important driver of tree mortality, as
shown by \citep{Senf2020} in Europe. However, we found that forest is
the type of landcover macroclass less affected by variation in drought
indices, being the most resistant landcover class to drought. Supporting
this is \citep{Fathi-Taperasht2022}, who asserts that Indian forests are
the most drought-resistant and recover rapidly. Similarly, the work of
\citep{Wu2024} who analyzed vegetation loss and recovery in response to
meteorological drought in the humid subtropical Pearl River basin in
China, indicates that forests showed higher drought resistance.
\citep{Xiao2023} using Vegetation Optical Depth (VOD), kNDVI, and EVI,
evaluates ecosystem resistance, showing that, compared to croplands,
ecosystems with a higher proportion of forests are more resilient to
severe droughts. They attribute the difference to a deeper rooting depth
of trees, a higher water storage capacity, and different water use
strategies between them \citep{Xiao2023}.

2.- Sobre los tipos de land cover más afectados por los indicadores de
sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur
(\citep{Miranda2023})

\citep{Fathi-Taperasht2022}: muestra que bosques son más resilentes a la
sequía en India.

\citep{Wu2024}: evalua la perdida de vegetación en diferentes tipos de
uso de suelo. Muestra que bosques son más resilentes y cultivos y
pastizales, los menos.

\citep{Jiang2020}: impacto de la sequía en landcover en China.
